On the power of conditional independence testing \\ under model-X

### 2022-05-10 arxiv-v4/crtpower.tex
Finalized EJS revision

Testing CI hypotheses in the MX framework has been the subject of active methodological research. The most popular methodology is MX knockoffs \cite{CetL16}. This method is based on the idea of constructing synthetic negative controls (knockoffs) for each predictor variable in a rigorous way that is based on the MX assumption; see Section~\ref{sec:knockoffs-overview} for a brief overview. Rapid progress has been made on the construction of knockoffs in various cases \cite{SetC17,Romano2019a,Bates2019,Huang2019} and on the application of this methodology to GWAS \cite{SetC17, SetS19}. The conditional randomization test (CRT) \cite{CetL16}, initially less popular than knockoffs due to its computational cost, is receiving renewed attention as computationally efficient variants are proposed, such as the holdout randomization test (HRT) \cite{Tansey2018}, the digital twin test \cite{Bates2020}, and the distilled CRT (dCRT) \cite{Liu2020}. We describe the CRT methodology next.

@@

\paragraph{A2: The prediction error of the machine learning method impacts the asymptotic efficiency of the CRT but not its consistency.}

It has been widely observed that the better the machine learning method approximates $\pry|\prx,\prz$, the higher power the MX method will have. We put this empirical knowledge on a theoretical foundation by expressing the asymptotic power of the CRT in terms of the prediction error of the underlying machine learning method (Section~\ref{sec:asymptotic-power}). In particular, we consider semiparametric alternatives of the form
\begin{equation}
	H_1: \mathcal L(\pry|\prx,\prz) = N(\prx^T\beta + g(\prz),\sigma^2).
	\label{parametric-alternative-intro}
\end{equation}
We analyze the power of a CRT variant that employs a separately trained estimator $\widehat g$ in an asymptotic regime where $d = \text{dim}(\prx)$ remains fixed while $p = \text{dim}(\prz)$ grows arbitrarily with the sample size $n$. We find that this test is consistent no matter what $\widehat g$ is used, while its asymptotic power against local alternatives $\beta_n = h/\sqrt{n}$ depends on the limiting mean-squared prediction error of $\widehat g$ (denoted $\mathcal E^2$) and the limiting expected variance $\mathbb E[\text{Var}[\prx | \prz]]$ (denoted $s^2$). For example, if $d = 1$, the
\begin{equation*}
	\begin{split}
		\text{CRT power converges to that of normal location test under alternative } N\smash{\left(\frac{hs}{\sqrt{\sigma^2 + \mathcal E^2}},1\right)}.
	\end{split}
\end{equation*}
This represents the first explicit quantification of the impact of machine learning prediction error on the power of an MX method. \\

@@

\paragraph{A resampling-free second-order approximation to the CRT is equivalent to the CRT and controls Type-I error under weaker assumptions.}

@@

In this section, we seek the most powerful MX CI test against a point alternative. To accomplish this, we observe that any (marginally) level $\alpha$ test must also have level $\alpha$ conditionally on $\sry, \srz$ (Section~\ref{sec:reduction-by-conditioning}). The latter conditioning step reduces the composite null to a point null. This reduction allows us to invoke the Neyman-Pearson lemma to find the most powerful test  (Section~\ref{sec:NP}). Proofs are deferred to Appendix~\ref{sec:proofs-sec2}.

@@

Proposition~\ref{prop:marginal-implies-conditional} allows us to reframe the MX CI problem as that of testing a null hypothesis with respect to the conditional law $\mathcal L(\srx|\sry,\srz)$, viewing only $\srx$ as random while fixing $\sry$ and $\srz$ at their observed values. It states that any level $\alpha$ test $\phi$, when viewed as a \textit{family} of hypothesis tests $\phi(\srx; y,z)$ indexed by $(\sfy,\sfz)$, has level $\alpha$ in the conditional testing problem for each $(\sfy, \sfz)$. This conditional perspective is useful because it reduces the composite null \eqref{conditional-independence} to a point null. Indeed, note that 
\begin{equation}
	\mathcal L \in \mathscr L_0^{\text{MX}}(f^*) \Longrightarrow \mathcal L(\srx = \sfx|\sry = \sfy, \srz = \sfz) = \prod_{i = 1}^n  f^*(\sfx_i | \sfz_i).
	\label{point-null}
\end{equation}
Therefore, $\mathcal L(\srx|\sry = \sfy, \srz = \sfz)$ equals a fixed product distribution for any null $\mathcal L$. This yields a conditional point null hypothesis. Note that the observations $\srx_i$ are independent \textit{but not identically distributed} due to the different conditioning events in~\eqref{point-null}.

@@

%The form of the optimal statistic $T^{\text{opt}}$ may be surprising, as it is a likelihood ratio with respect to $\sry|\srx,\srz$ rather than $\srx|\sry,\srz$. In fact, we see in the proof (Appendix~\ref{sec:proofs-sec2}) that these two likelihood ratios are equivalent, with the former yielding a more explicit expression. Note also that the denominator of $T^{\text{opt}}$ can be omitted if desired, since it is not a function of $\srx$. We keep it for a more direct parallel with the model-$\sry|\srx$ case. % Though this paper is primarily concerned with the CRT, we also obtain a parallel result about MX knockoffs in Section~\ref{sec:knockoffs}. 
The proof of Theorem~\ref{prop:crt-optimality} is based on the reduction in Section~\ref{sec:reduction-by-conditioning} of the composite null to a point null by conditioning, followed by the Neyman-Pearson lemma. Note that the likelihood ratio in the model $\mathcal L(\prx|\pry, \prz)$ reduces to the likelihood in the model $\mathcal L(\pry|\prx, \prz)$ up to constant factors; see derivation~\eqref{eq:likelihood-ratio-derivation}. This argument has similar flavor to the theory of unbiased testing (see Lehmann and Romano \cite[Chapter 4]{TSH}), where uniformly most powerful unbiased tests can be found by conditioning on sufficient statistics for nuisance parameters. Our result is also analogous to but different from Lehmann's derivation of the most powerful permutation tests using conditioning followed by the Neyman-Pearson lemma, in randomization-based causal inference (see the rejoinder of Rosenbaum's 2002 discussion paper \cite{Rosenbaum2002}, Section 5.10 of Lehmann (1986), now Lehmann and Romano \cite[Section 5.9]{TSH}).

@@

In Section~\ref{sec:power}, we saw how to construct the optimal test against point alternatives specified by $\bar f_{\pry|\prx,\prz}$. In practice, of course we do not have access to this distribution, so we usually estimate it via a statistical machine learning procedure. The goal of this section and the next is to quantitatively assess the power of the CRT as a function of the prediction error of this machine learning procedure. Specifically, we consider  the power of a specific instance of the CRT (the \textit{distilled CRT} \cite{Liu2020})  against a set of semiparametric alternatives (Section~\ref{sec:asymptotic-setup}). We prepare to assess the power of this test by showing its asymptotic equivalence to the simpler-to-analyze \textit{MX(2) $F$-test} (Section~\ref{sec:second-order-approx}), which is of independent interest due to its closed form and weaker assumptions (Section~\ref{sec:mx2-f-test}). We examine the finite-sample Type-I error control of the MX(2) $F$-test in numerical simulations (Section~\ref{sec:simulations}) and put this section's results into perspective (Section~\ref{sec:comparison-to-existing-results-3}) before moving on to stating the desired power results in the next section (Section~\ref{sec:asymptotic-power}).

@@

Of course, inference based on $T_n^{\text{opt}}$ is infeasible because the function $g'_n$ is unknown in practice. Suppose we have learned an estimate $\widehat g_n$ of this function, possibly in-sample. Then, the derivation~\eqref{eq:optimal-semiparametric} motivates us to base inference on the sample covariance between $\prx$ and $\pry$ after adjusting for $\prz$:
\begin{equation}
	\widehat \rho_n(\srx, \sry, \srz) \equiv \frac1n\sum_{i = 1}^n (\sry_{i} - \widehat g_n(\srz_{i}))(\srx_{i} - \mu_n(\srz_i)).
	\label{rho-hat}
\end{equation}
Consider first the case $d = 1$. The CRT rejecting for large values of $|\widehat \rho_n|$ is an instance of the dCRT \cite{Liu2020}. The idea of the dCRT (Algorithm~\ref{alg:dCRT}) is to \textit{distill}---usually via machine learning---the information from the high-dimensional $\srz \in \mathbb R^{p \times n}$ about $\srx$ and $\sry$ into a low-dimensional summary $D \in \mathbb R^{q \times n}$,  where $q \ll p$. This is accomplished using a \textit{distillation function} $d: (\sry, \srz) \mapsto D$. Then, the CRT is applied using a test statistic of the form $T_n(\srx, \sry, \srz) \equiv T_n^d(\srx, \sry, D) = T_n^d(\srx, \sry, d(\sry, \srz))$. For example, the CRT based on the statistic $\widehat \rho_n$~\eqref{rho-hat} can be expressed as the dCRT with distillation function $d_i(\sry, \srz) = (\widehat g_n(\srz_i), \mu_n(Z_i))$, where $\widehat g_n$ is learned in-sample on $(\sry, \srz)$.

\begin{center}
	\begin{minipage}{\linewidth}
		\begin{algorithm}[H]
			%			\SetAlgorithmName{Procedure}{}\; %last arg is the title of listing table		
			\nonl  \textbf{Input:} $\{(\srx_i, \sry_i, \srz_i)\}_{i=1}^n$, distribution  $f^*_{\prx|\prz}$, distillation function $d$, test statistic $T_n^d$, number of resamples $B$\\
			Distill information in $\srz$ about $\srx$ and $\sry$ into $D \equiv d(\sry, \srz)$\;
			\For{$b = 1, 2, \dots, B$}{
				Resample $\srxk_i^{(b)} \overset{\text{ind}}\sim f^*_{\prx|\prz = \srz_i},\ i = 1, \dots, n$\;
			}
			Compute $\widehat p \equiv \frac{1}{B+1}\sum_{b = 1}^B \mathbbm 1(T_n^d(\srxk^{(b)}, \sry, D) \geq T_n^d(\srx, \sry, D))$.\\
			\nonl \textbf{Output:} dCRT $p$-value $\widehat p$.\\
			\nonl \textbf{Computational cost:} One $p$-dimensional model fit, and drawing $B$ resamples.
			\caption{\bf The distilled conditional randomization test}
			\label{alg:dCRT}
		\end{algorithm}
	\end{minipage}
\end{center}

@@

Informally, this theorem suggests that the CRT resampling distribution of $T_n(\srx, \sry, \srz)$ converges to $\chi^2_d$, which is the resampling distribution of this test statistic under a normal $\mathcal L_n(\prx|\prz)$. Note that the resulting equivalence~\eqref{eq:asymptotic-equivalence} holds for the specific instance of the CRT based on the statistic $T_n$ defined in via equations~\eqref{u-hat} and~\eqref{t-n}, though other kinds of test statistics may lead to similar large-sample behavior. While Theorem~\ref{thm:equivalence} is stated for $\widehat g_n$ trained out of sample, we conjecture that it continues to hold when $\widehat g_n$ is fit in sample, as in the original dCRT construction \cite{Liu2020}.

@@

The moment assumptions can be relaxed for pointwise error control (less desirable), but are unavoidable for uniform type-1 error control as stated in the corollary. More importantly, we conjecture that the MX(2) $F$-test continues to have asymptotic Type-I error control even if $\widehat g_n$ is fit in sample. This conjecture is supported by the results of a simulation study, described next.

@@

Recalling that $\chi^2_d(\lambda)$ denotes the noncentral chi-square distribution with $d$ degrees of freedom and non-centrality parameter $\lambda$ and $c_{d,1-\alpha}$ denotes the $1-\alpha$ quantile of $\chi^2_d$, the second part of Theorem~\ref{thm:power} states that the dCRT has power equal to that of a $\chi^2$ test of a multivariate normal random vector having mean zero under the alternative $N((\sigma^2I_d +\mathcal E^2)^{-1/2}\overline \Sigma^{1/2} h, I_d)$. This result establishes a direct link between the estimation error in $\widehat g_n$ and the power of the CRT against local alternatives. In particular, the mean-squared error term $\mathcal E^2$ contributes additively to the irreducible error term $\sigma^2 I_d$. We can gain intuition for this result by considering the regression model
\begin{equation}
\begin{split}
\pry - \widehat g_n(\prz) &= (\prx - \mu_n(\prz))^T\beta_n + (g'_n(\prz) - \widehat g_n(\prz) + \bm \eps)
\label{eq:regression-model}
\end{split}
\end{equation}
obtained from the semiparametric model~\eqref{eq:linearity} by subtracting $\widehat g_n(\prz)$ from both sides. The test statistic $T_n$ is based on the quantity $\widehat \rho_n$ defined in equation~\eqref{rho-hat}, which can be viewed as an unnormalized version of the fitted regression coefficients of $\sry - \widehat g_n(\srz)$ on $\srx - \mu_n(\srz)$. The term $g'_n(\prz) - \widehat g_n(\prz)$ in the regression model~\eqref{eq:regression-model} contributes additively to the residual error term, so in a traditional regression analysis we would expect the power of the test to depend on the variance of this error term. In fact, standard large-sample OLS theory (see e.g. Section 2.3 of Hayashi's book \cite{Hayashi2000}) states that the power against local alternatives of the $F$-test in the regression model~\eqref{eq:regression-model} is exactly the same as that of the dCRT (and MX(2) $F$-test) stated in equation~\eqref{eq:main-conclusion}. Of course, the usual $F$-test applied to the regression~\eqref{eq:regression-model} relies on the validity of this model while the dCRT and MX(2) $F$-test instead rely on knowledge of $\text{Var}[\prx|\prz]$. Note that \cite{Wang2020b} also find the power of an MX test and a classical OLS test to have the same power (see their Appendix F).

@@

%The optimal test statistic~\eqref{log-likelihood-ratio-knockoffs} has a stunning simplicity: it is the likelihood of the response given the observed variables. 
The reader observes that the optimal test statistic is not a function of the knockoff variables or of the index $j$, which may at first seem paradoxical. Recall from the definition~\eqref{one-bit-pvalue}, however, that the one-bit $p$-value compares the test statistic on the original augmented design $[\srx, \srxk]$ and its swapped version $[\srx, \srxk]_{\textnormal{swap}(j)}$. Therefore, the optimal one-bit $p$-value checks whether the original $j$th variable $\srx_{\bullet, j}$ fits with the rest of the data better than does its knockoff $\srxk_{\bullet, j}$. Therefore, the optimal one-bit $p$-value is in fact a function of the knockoffs as well as the index $j$. A simple way of operationalizing Theorem~\ref{prop:knockoff-optimality} is to fit a model $\widehat f(\pry|\prx)$ based on $([\srx, \srxk], \sry)$ in any way that treats original variables and knockoffs symmetrically, and then defining $T_j([\srx, \srxk], \sry) \equiv \widehat f(\sry|\srx)$. The above result continues to hold when $\prx_j$ is a \textit{group} of variables, giving a clean way to combine evidence across multiple variables. A conditional version of the optimality statement~\eqref{unconditional-knockoff-optimality} holds; see equation~\eqref{knockoff-conditional-optimality} in the appendix.

Theorem~\ref{prop:knockoff-optimality} requires that ties occur with probability zero~\eqref{nondegeneracy-assumption}. Proposition~\ref{prop:nondegeneracy-knockoffs} below states that this nondegeneracy condition holds if either $\pry|\prx$ or $\prx_j|\prx_{-j}, \prxk$ has a continuous distribution.
\begin{proposition}\label{prop:nondegeneracy-knockoffs}
	Suppose $\bar{\mathcal L}(\pry|\prx) = g_{\bm\eta}$, where $\bm \eta = \prx_j \beta_j + f_{-j}(\prx_{-j})$ and $g_\eta$ is a one-dimensional exponential family with natural parameter $\eta$ and strictly convex, continuous log partition function $\psi$. Suppose also that $\prx_j, \beta_j \in \mathbb R$, with $\beta_j \neq 0$. The nondegeneracy condition~\eqref{nondegeneracy-assumption} holds if either 
	\begin{enumerate}
		\item $\prx_{j}|\prx_{-j}, \prxk$ has a density for each $\prx_{-j}, \prxk$, or
		\item $g_\eta$ has a density,
	\end{enumerate}
	where the densities are with respect to the Lebesgue measure.
\end{proposition}

>>>>

Testing CI hypotheses in the MX framework has been the subject of active methodological research. The most popular methodology is MX knockoffs \cite{CetL16}. This method is based on the idea of constructing synthetic negative controls (knockoffs) for each predictor variable in a rigorous way that is based on the MX assumption; see Section~\ref{sec:knockoffs-overview} for a brief overview. Rapid progress has been made on the construction of knockoffs in various cases \cite{SetC17,Romano2019a,Bates2019,Huang2019} and on the application of this methodology to GWAS \cite{SetC17, SetS19}. The conditional randomization test (CRT) \cite{CetL16}, initially less popular than knockoffs due to its computational cost, is receiving renewed attention as computationally efficient variants are proposed, such as the holdout randomization test (HRT) \cite{Tansey2018}, the digital twin test \cite{Bates2020}, and the distilled CRT (dCRT) \cite{Liu2020}. The dCRT in particular is a promising methodology because it combines good power and computational speed; we focus on this variant of the CRT is Sections~\ref{sec:weakening} and~\ref{sec:asymptotic-power} of this paper. We introduce the general CRT methodology next, while deferring the introduction of the dCRT to Section~\ref{sec:weakening}.

@@

\paragraph{A2: The prediction error of the machine learning method impacts the asymptotic efficiency of the dCRT but not its consistency.}

It has been widely observed that the better the machine learning method approximates $\pry|\prx,\prz$, the higher power the MX method will have. We put this empirical knowledge on a theoretical foundation by expressing the asymptotic power of the dCRT in terms of the prediction error of the underlying machine learning method (Section~\ref{sec:asymptotic-power}). In particular, we consider semiparametric alternatives of the form
\begin{equation}
	H_1: \mathcal L(\pry|\prx,\prz) = N(\prx^T\beta + g(\prz),\sigma^2).
	\label{parametric-alternative-intro}
\end{equation}
We analyze the power of a dCRT variant that employs a separately trained estimator $\widehat g$ in an asymptotic regime where $d = \text{dim}(\prx)$ remains fixed while $p = \text{dim}(\prz)$ grows arbitrarily with the sample size $n$. We find that this test is consistent no matter what $\widehat g$ is used, while its asymptotic power against local alternatives $\beta_n = h/\sqrt{n}$ depends on the limiting mean-squared prediction error of $\widehat g$ (denoted $\mathcal E^2$) and the limiting expected variance $\mathbb E[\text{Var}[\prx | \prz]]$ (denoted $s^2$). For example, if $d = 1$, the
\begin{equation*}
	\begin{split}
		\text{dCRT power converges to that of normal location test under alternative } N\smash{\left(\frac{hs}{\sqrt{\sigma^2 + \mathcal E^2}},1\right)}.
	\end{split}
\end{equation*}
This represents the first explicit quantification of the impact of machine learning prediction error on the power of an MX method. \\

@@

\paragraph{A resampling-free second-order approximation to the dCRT is equivalent to the dCRT and controls Type-I error under weaker assumptions.}

@@

In this section, we seek the most powerful MX CI test against a point alternative. To accomplish this, we observe that any (marginally) level $\alpha$ test must also have level $\alpha$ conditionally on $\sry, \srz$ (Section~\ref{sec:reduction-by-conditioning}). The latter conditioning step reduces the composite null to a point null. This reduction allows us to invoke the Neyman-Pearson lemma to find the most powerful test  (Section~\ref{sec:NP}). Proofs are deferred to the appendix.

@@

Proposition~\ref{prop:marginal-implies-conditional}, whose proof is given in Appendix~\ref{sec:proofs-sec2}, allows us to reframe the MX CI problem as that of testing a null hypothesis with respect to the conditional law $\mathcal L(\srx|\sry,\srz)$, viewing only $\srx$ as random while fixing $\sry$ and $\srz$ at their observed values. It states that any level $\alpha$ test $\phi$, when viewed as a \textit{family} of hypothesis tests $\phi(\srx; y,z)$ indexed by $(\sfy,\sfz)$, has level $\alpha$ in the conditional testing problem for each $(\sfy, \sfz)$. This conditional perspective is useful because it reduces the composite null \eqref{conditional-independence} to a point null. Indeed, note that 
\begin{equation}
	\mathcal L \in \mathscr L_0^{\text{MX}}(f^*) \Longrightarrow \mathcal L(\srx = \sfx|\sry = \sfy, \srz = \sfz) = \prod_{i = 1}^n  f^*(\sfx_i | \sfz_i).
	\label{point-null}
\end{equation}
Therefore, $\mathcal L(\srx|\sry = \sfy, \srz = \sfz)$ equals a fixed product distribution for any null $\mathcal L$. This yields a conditional point null hypothesis. Note that the observations $\srx_i$ are independent \textit{but not identically distributed} due to the different conditioning events in~\eqref{point-null}.

@@

%The form of the optimal statistic $T^{\text{opt}}$ may be surprising, as it is a likelihood ratio with respect to $\sry|\srx,\srz$ rather than $\srx|\sry,\srz$. In fact, we see in the proof (Appendix~\ref{sec:proofs-sec2}) that these two likelihood ratios are equivalent, with the former yielding a more explicit expression. Note also that the denominator of $T^{\text{opt}}$ can be omitted if desired, since it is not a function of $\srx$. We keep it for a more direct parallel with the model-$\sry|\srx$ case. % Though this paper is primarily concerned with the CRT, we also obtain a parallel result about MX knockoffs in Section~\ref{sec:knockoffs}. 
The proof of Theorem~\ref{prop:crt-optimality} (Appendix~\ref{sec:proofs-sec2}) is based on the reduction in Section~\ref{sec:reduction-by-conditioning} of the composite null to a point null by conditioning, followed by the Neyman-Pearson lemma. Note that the likelihood ratio in the model $\mathcal L(\prx|\pry, \prz)$ reduces to the likelihood in the model $\mathcal L(\pry|\prx, \prz)$ up to constant factors; see derivation~\eqref{eq:likelihood-ratio-derivation}. This argument has similar flavor to the theory of unbiased testing (see Lehmann and Romano \cite[Chapter 4]{TSH}), where uniformly most powerful unbiased tests can be found by conditioning on sufficient statistics for nuisance parameters. Our result is also analogous to but different from Lehmann's derivation of the most powerful permutation tests using conditioning followed by the Neyman-Pearson lemma, in randomization-based causal inference (see the rejoinder of Rosenbaum's 2002 discussion paper \cite{Rosenbaum2002}, Section 5.10 of Lehmann (1986), now Lehmann and Romano \cite[Section 5.9]{TSH}).

@@

In Section~\ref{sec:power}, we saw how to construct the optimal test against point alternatives specified by $\bar f_{\pry|\prx,\prz}$. In practice, of course we do not have access to this distribution, so we usually estimate it via a statistical machine learning procedure. The goal of this section and the next is to quantitatively assess the power of the CRT as a function of the prediction error of this machine learning procedure. Specifically, we consider  the power of a specific instance of the CRT (the \textit{distilled CRT (dCRT)} \cite{Liu2020})  against a set of semiparametric alternatives (Section~\ref{sec:asymptotic-setup}). We prepare to assess the power of this test by showing its asymptotic equivalence to the simpler-to-analyze \textit{MX(2) $F$-test} (Section~\ref{sec:second-order-approx}), which is of independent interest due to its closed form and weaker assumptions (Section~\ref{sec:mx2-f-test}). We examine the finite-sample Type-I error control of the MX(2) $F$-test in numerical simulations (Section~\ref{sec:simulations}) and put this section's results into perspective (Section~\ref{sec:comparison-to-existing-results-3}) before moving on to stating the desired power results in the next section (Section~\ref{sec:asymptotic-power}).

@@

Of course, inference based on $T_n^{\text{opt}}$ is infeasible because the function $g'_n$ is unknown in practice. Suppose we have learned an estimate $\widehat g_n$ of this function, possibly in-sample. Then, the derivation~\eqref{eq:optimal-semiparametric} motivates us to base inference on the sample covariance between $\prx$ and $\pry$ after adjusting for $\prz$:
\begin{equation}
	\widehat \rho_n(\srx, \sry, \srz) \equiv \frac1n\sum_{i = 1}^n (\sry_{i} - \widehat g_n(\srz_{i}))(\srx_{i} - \mu_n(\srz_i)).
	\label{rho-hat}
\end{equation}
Consider first the case $d = 1$. The CRT rejecting for large values of $|\widehat \rho_n|$ is an instance of the dCRT \cite{Liu2020}. The idea of the dCRT (Algorithm~\ref{alg:dCRT}) is to \textit{distill}---usually via a machine learning regression method---the information from the high-dimensional $\srz \in \mathbb R^{p \times n}$ about $\srx$ and $\sry$ into a low-dimensional summary $D \in \mathbb R^{q \times n}$,  where $q \ll p$. This is accomplished using a \textit{distillation function} $d: (\sry, \srz) \mapsto D$. Then, the CRT is applied using a test statistic of the form $T_n(\srx, \sry, \srz) \equiv T_n^d(\srx, \sry, D) = T_n^d(\srx, \sry, d(\sry, \srz))$. For example, the CRT based on the statistic $\widehat \rho_n$~\eqref{rho-hat} can be expressed as the dCRT with distillation function $d_i(\sry, \srz) = (\widehat g_n(\srz_i), \mu_n(Z_i))$, where $\widehat g_n$ is learned in-sample on $(\sry, \srz)$.

\begin{center}
	\begin{minipage}{\linewidth}
		\begin{algorithm}[H]
			%			\SetAlgorithmName{Procedure}{}\; %last arg is the title of listing table		
			\nonl  \textbf{Input:} $\{(\srx_i, \sry_i, \srz_i)\}_{i=1}^n$, distribution  $f^*_{\prx|\prz}$, distillation function $d$, test statistic $T_n^d$, number of resamples $B$\\
			Distill information in $\srz$ about $\srx$ and $\sry$ into $D \equiv d(\sry, \srz)$\;
			\For{$b = 1, 2, \dots, B$}{
				Resample $\srxk_i^{(b)} \overset{\text{ind}}\sim f^*_{\prx|\prz = \srz_i},\ i = 1, \dots, n$\;
			}
			Compute $\widehat p \equiv \frac{1}{B+1}\sum_{b = 1}^B \mathbbm 1(T_n^d(\srxk^{(b)}, \sry, D) \geq T_n^d(\srx, \sry, D))$.\\
			\nonl \textbf{Output:} dCRT $p$-value $\widehat p$.\\
			\nonl \textbf{Computational cost:} One $p$-dimensional model fit, and drawing $B$ resamples.
			\caption{\bf The distilled conditional randomization test (dCRT)}
			\label{alg:dCRT}
		\end{algorithm}
	\end{minipage}
\end{center}

@@

Informally, this theorem (proved in Appendix~\ref{sec:proofs-sec45}) suggests that the CRT resampling distribution of $T_n(\srx, \sry, \srz)$ converges to $\chi^2_d$, which is the resampling distribution of this test statistic under a normal $\mathcal L_n(\prx|\prz)$. Note that the resulting equivalence~\eqref{eq:asymptotic-equivalence} holds for the specific instance of the CRT based on the statistic $T_n$ defined in via equations~\eqref{u-hat} and~\eqref{t-n}, though other kinds of test statistics may lead to similar large-sample behavior. While Theorem~\ref{thm:equivalence} is stated for $\widehat g_n$ trained out of sample, we conjecture that it continues to hold when $\widehat g_n$ is fit in sample, as in the original dCRT construction \cite{Liu2020}.

@@

See Appendix~\ref{sec:proofs-sec45} for a proof of this theorem. The moment assumptions can be relaxed for pointwise error control (less desirable), but are unavoidable for uniform type-1 error control as stated in the corollary. More importantly, we conjecture that the MX(2) $F$-test continues to have asymptotic Type-I error control even if $\widehat g_n$ is fit in sample. This conjecture is supported by the results of a simulation study, described next.

@@

This theorem is proved in Appendix~\ref{sec:proofs-sec45}. Recalling that $\chi^2_d(\lambda)$ denotes the noncentral chi-square distribution with $d$ degrees of freedom and non-centrality parameter $\lambda$ and $c_{d,1-\alpha}$ denotes the $1-\alpha$ quantile of $\chi^2_d$, the second part of Theorem~\ref{thm:power} states that the dCRT has power equal to that of a $\chi^2$ test of a multivariate normal random vector having mean zero under the alternative $N((\sigma^2I_d +\mathcal E^2)^{-1/2}\overline \Sigma^{1/2} h, I_d)$. This result establishes a direct link between the estimation error in $\widehat g_n$ and the power of the CRT against local alternatives. In particular, the mean-squared error term $\mathcal E^2$ contributes additively to the irreducible error term $\sigma^2 I_d$. We can gain intuition for this result by considering the regression model
\begin{equation}
\begin{split}
\pry - \widehat g_n(\prz) &= (\prx - \mu_n(\prz))^T\beta_n + (g'_n(\prz) - \widehat g_n(\prz) + \bm \eps)
\label{eq:regression-model}
\end{split}
\end{equation}
obtained from the semiparametric model~\eqref{eq:linearity} by subtracting $\widehat g_n(\prz)$ from both sides. The test statistic $T_n$ is based on the quantity $\widehat \rho_n$ defined in equation~\eqref{rho-hat}, which can be viewed as an unnormalized version of the fitted regression coefficients of $\sry - \widehat g_n(\srz)$ on $\srx - \mu_n(\srz)$. The term $g'_n(\prz) - \widehat g_n(\prz)$ in the regression model~\eqref{eq:regression-model} contributes additively to the residual error term, so in a traditional regression analysis we would expect the power of the test to depend on the variance of this error term. In fact, standard large-sample OLS theory (see e.g. Section 2.3 of Hayashi's book \cite{Hayashi2000}) states that the power against local alternatives of the $F$-test in the regression model~\eqref{eq:regression-model} is exactly the same as that of the dCRT (and MX(2) $F$-test) stated in equation~\eqref{eq:main-conclusion}. Of course, the usual $F$-test applied to the regression~\eqref{eq:regression-model} relies on the validity of this model while the dCRT and MX(2) $F$-test instead rely on knowledge of $\text{Var}[\prx|\prz]$. Note that \cite{Wang2020b} also find the power of an MX test and a classical OLS test to have the same power (see their Appendix F).

@@

%The optimal test statistic~\eqref{log-likelihood-ratio-knockoffs} has a stunning simplicity: it is the likelihood of the response given the observed variables. 
This theorem is proved in Appendix~\ref{sec:knockoffs-proofs}. The reader observes that the optimal test statistic is not a function of the knockoff variables or of the index $j$, which may at first seem paradoxical. Recall from the definition~\eqref{one-bit-pvalue}, however, that the one-bit $p$-value compares the test statistic on the original augmented design $[\srx, \srxk]$ and its swapped version $[\srx, \srxk]_{\textnormal{swap}(j)}$. Therefore, the optimal one-bit $p$-value checks whether the original $j$th variable $\srx_{\bullet, j}$ fits with the rest of the data better than does its knockoff $\srxk_{\bullet, j}$. Therefore, the optimal one-bit $p$-value is in fact a function of the knockoffs as well as the index $j$. A simple way of operationalizing Theorem~\ref{prop:knockoff-optimality} is to fit a model $\widehat f(\pry|\prx)$ based on $([\srx, \srxk], \sry)$ in any way that treats original variables and knockoffs symmetrically, and then defining $T_j([\srx, \srxk], \sry) \equiv \widehat f(\sry|\srx)$. The above result continues to hold when $\prx_j$ is a \textit{group} of variables, giving a clean way to combine evidence across multiple variables. A conditional version of the optimality statement~\eqref{unconditional-knockoff-optimality} holds; see equation~\eqref{knockoff-conditional-optimality} in the appendix.

Theorem~\ref{prop:knockoff-optimality} requires that ties occur with probability zero~\eqref{nondegeneracy-assumption}. Proposition~\ref{prop:nondegeneracy-knockoffs} below (proved in Appendix~\ref{sec:proofs-sec45}) states that this nondegeneracy condition holds if either $\pry|\prx$ or $\prx_j|\prx_{-j}, \prxk$ has a continuous distribution.
\begin{proposition}\label{prop:nondegeneracy-knockoffs}
	Suppose $\bar{\mathcal L}(\pry|\prx) = g_{\bm\eta}$, where $\bm \eta = \prx_j \beta_j + f_{-j}(\prx_{-j})$ and $g_\eta$ is a one-dimensional exponential family with natural parameter $\eta$ and strictly convex, continuous log partition function $\psi$. Suppose also that $\prx_j, \beta_j \in \mathbb R$, with $\beta_j \neq 0$. The nondegeneracy condition~\eqref{nondegeneracy-assumption} holds if either 
	\begin{enumerate}
		\item $\prx_{j}|\prx_{-j}, \prxk$ has a density for each $\prx_{-j}, \prxk$, or
		\item $g_\eta$ has a density,
	\end{enumerate}
	where the densities are with respect to the Lebesgue measure.
\end{proposition}
